---
title: "EDA"
output: html_document
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Importing libraries
```{r include=FALSE}
library(ggplot2)
library(knitr)
library(tidyverse)
library(dplyr)
library(kableExtra)
library(gridExtra)
library(ggplot2)
```

```{r load-data}
load('./30103-0001-Data.rda')
df <- da30103.0001
#names(df)
#str(df)
```

Data Cleaning
```{r cols-to-keep}
keeps <- c('RESPONDENT_YRSED',# Approx Respondent yrs education based on ppeduc, first demographic survey
           'PARTNER_YRSED',   #Partner yrs education based on q10
           'PPINCIMP',        #Binned - total combined household income
           'PPWORK',          #employment
           'QFLAG',           # does person have a partner?
           'PAPGLB_STATUS',   #(gay, lesbian, or bisexual)
           'PPMARIT',         # marital status
           'PPGENDER',        # gender
           'Q4',              #( is partner female or male) 
           'Q5',              #(is partner same gender)
           'Q23',             # between you and [partner_name], who earned more income in 2008
           'PPETHM',          # race/ethnicity hispanic/latino
           'Q6B',             # race of partner
           'RESPONDENT_RACE',
           'PARTNER_RACE',
           'RELATIONSHIP_QUALITY', # relationship quality, based on q34, higher number is better
           'PPGENDER', #gender
           'PPREG4', #based on state of residence
           'PPREG9', #based on state of residence
           'PPWORK', #current employment status
           'SAME_SEX_COUPLE', #same sex couple
           'PPAGE', #age at the time 
           'PAPRELIGION', #religion
           'Q7B', #partner's religion
           'PPQ14ARACE', #race/ethnicity
           'PPEDUC', #Education (highest degree received)
           'PPEDUCAT', #Education categorical
           'PPHOUSE', #Housing type
           'PPRENT', #Ownership status
           'PPMARIT', #marital status
           'PPPARTYID3', #Political Party
           'Q20', #cohabitating
           'Q19' #cohabit
          )

# store as new df
df <- df[keeps]

# only keep the partnered people
df <- df[df$QFLAG=='(1) partnered',]

```

### Data Transformation
```{r add-col}
# add a new column to track difference in education
df['diff_in_education'] = df$RESPONDENT_YRSED - df$PARTNER_YRSED

# add boolean to track difference in race
df['diff_in_race'] = ifelse(df$RESPONDENT_RACE == df$PARTNER_RACE, 0, 1)

# add boolean to track difference in income
df['diff_in_income'] = ifelse(df$Q23 == '(2) we earned about the same amount', 0, 1)

#add Relationship quality
df$relationship_val <- as.numeric(substr(df$RELATIONSHIP_QUALITY, 2, 2))

#add PPWORK
df$work <- as.numeric((substr(df$PPWORK, 2,2)))
#head(df$work)

#PPAGE - numeric
df$PPAGE <- as.numeric(df$PPAGE)

#Convert PPINCIMP to numeric
#df$PPINCIMP <- as.numeric(gsub('[$,]', '', df$PPINCIMP))
```

## Exploratory Data Analysis 

1. Relationship Quality -- new column relationship_val

```{r quality}
glimpse(df$relationship_val)

(summary(df$relationship_val))

ggplot(df, aes(relationship_val)) + geom_bar() + theme_bw() + xlab("Relationship Quality") + ylab("Count") 

NA %in% df$RELATIONSHIP_QUALITY
sum(is.na(df$RELATIONSHIP_QUALITY))

#There's 13 NA values for relationship quality - removing them
df <- df[is.na(df$RELATIONSHIP_QUALITY)==FALSE,]

ggplot(df, aes(relationship_val)) + geom_bar() + theme_bw() + xlab("Relationship Quality") + ylab("Count") + scale_fill_discrete(name = "Title", labels = c("A","B","C","D","E")) + theme(legend.position = "top")
```

Most subject quantify their relationship status as excellent.

2. Current Employment Status

```{r employment}
kable(summary(df$PPWORK)) %>% kable_styling(font_size = 14)

ggplot(df, aes(PPWORK)) + geom_bar() + coord_flip() + theme_bw() + ylab("Count") + theme(legend.position = "bottom")

```

Most of the participants are working as a paid employee.

3. GLB Status

```{r glb}
kable(summary(df$PAPGLB_STATUS)) %>% kable_styling(font_size = 14)


ggplot(df, aes(PAPGLB_STATUS)) + geom_bar() + theme_bw() + xlab("GLB Status") + ylab("Count") + theme(legend.position = "bottom")

```

Approximately, 1/5 of the subjects identify as gay/lesbian/bi.

4. Same Sex couple

```{r same-sex}

kable(summary(df$SAME_SEX_COUPLE)) %>% kable_styling(font_size = 14)

ggplot(df, aes(SAME_SEX_COUPLE)) + geom_bar() + theme_bw() + xlab("Same Sex Couple") + ylab("Count") + theme(legend.position = "bottom")

```

There are 474 same sex couples (approximately 15%)

5. Age

```{r age}
summary(df$PPAGE)
ggplot(df, aes(PPAGE)) + geom_bar() + theme_bw() + xlab("Age")
```

Age ranges from 19 to 95, and is slightly right skewed with a median age of 45.


6. Gender

```{r gender}
kable(summary(df$Q4)) %>% kable_styling(font_size = 14)
ggplot(df, aes(Q4)) + geom_bar() + theme_bw() + xlab("Gender")
```

Approximately even ratio of males to females.


7. Comparing Religion

```{r religion}
kable(summary(df$PAPRELIGION)) %>% kable_styling(font_size = 14)

p1 <- ggplot(df, aes(PAPRELIGION)) + 
  geom_bar() + theme_bw() + 
  coord_flip() + ggtitle("Religion") +
  xlab("Subject's Religion")

p2 <- ggplot(df, aes(Q7B)) + 
  geom_bar() + theme_bw() + 
  coord_flip() + 
  ggtitle("Partner's Religion") + 
  xlab("Partner's Religion")

grid.arrange(p1,p2, ncol = 1)

```

8. Race/Ethnicity Distribution

```{r race}
kable(summary(df$PPETHM)) %>% kable_styling()

ggplot(df, aes(PPETHM)) + geom_bar() + theme_bw() + xlab("Race") + coord_flip()
```

A large majority of the subjects are white - likely over-represented.

9. Education Distribution

```{r education} 
kable(summary(df$PPEDUC)) %>% kable_styling()
kable(summary(df$PPEDUCAT)) %>% kable_styling()

p1 <- ggplot(df, aes(PPEDUC)) + geom_bar() + theme_bw() + xlab("Education") + coord_flip()
p2 <- ggplot(df, aes(PPEDUCAT)) + geom_bar() + theme_bw() + xlab("Education") + coord_flip()

grid.arrange(p1, p2, ncol = 1)
```

A majority of the subjects have a Bachelor's degree or higher.

10. Household Income Distribution

```{r income}
summary(df$PPINCIMP)

ggplot(df, aes(PPINCIMP)) + geom_bar() + theme_bw() + xlab("Household Income") + coord_flip()
```

The distribution is close to normal with some outliers at the high end.

11. Housing Type

```{r housing}
kable(summary(df$PPHOUSE)) %>% kable_styling()

ggplot(df, aes(PPHOUSE)) + geom_bar() + theme_bw() + xlab("Household Income") + coord_flip()
```

Most of the subjects live in a one family house.

12. Rent

```{r rent}
kable(summary(df$PPRENT)) %>% kable_styling()

ggplot(df, aes(PPRENT)) + geom_bar() + theme_bw() + xlab("Household Income") + coord_flip()
```

Most of the subjects own the house they live in.

13. Marital Status

```{r marital-status}
kable(summary(df$PPMARIT)) %>% kable_styling()

ggplot(df, aes(PPMARIT)) + geom_bar() + theme_bw() + xlab("Household Income") + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Most of the subjects are married.

14. Political Party Affiliation

```{r political}
kable(summary(df$PPPARTYID3)) %>% kable_styling()

ggplot(df, aes(PPPARTYID3)) + geom_bar() + theme_bw() + xlab("Household Income") 
```

Approximately 3:2 ratio of Democrats to Republicans with a small number of independents.

15. Cohabitating Couples

```{r}
kable(summary(df$Q19)) %>% kable_styling()

ggplot(df, aes(Q19)) + geom_bar() + theme_bw() + xlab("Living with Partner") 

```
2539 couples (79.6%) live with their partners.

16. Who Earns More?

```{r earning}
 
kable(summary(df$Q23)) %>% kable_styling(font_size = 14)

df %>% mutate(Q23 = fct_relevel(Q23, 
                           "We earned about the same amount",
                           "partner earned more", 
                           "I earned more")) %>% 
  
  mutate(Q23 = fct_recode(Q23,"We earned about the same" = "We earned about the same amount", "My partner earned more" = "[Partner Name] earned more")) %>% 
  
  # Removed those who refused from the dataset as well as those who reported that their partner was not working for pay -- a small total that do not contribute to any major trends
  
  filter(!is.na(Q23), 
         Q23 != "Refused", 
         Q23 != "[Partner Name] was not working for pay") %>%  # Create bar chart based on responses for Q23 (respondent's pay versus partner's pay)
  
  ggplot(aes(x = Q23, fill = PPGENDER)) + 
  
  geom_bar(show.legend=FALSE) + 
  
  facet_wrap(~PPGENDER) + 
  
  coord_flip() + 
  
  theme_bw() + 
  
  scale_fill_manual(values = c("gray34", "gray34")) +
  
  labs(
    title = "Earnings among Partners in the US", 
    subtitle = "Women were more likely to report making less than their partner", 
    caption = "Source: HCMST 2017", 
    x = NULL, y = NULL)
```

