---
title: "education"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
rm(list = ls())
df = read.csv("cleaned_df.csv")
df$relationship_val <- as.numeric(substr(df$RELATIONSHIP_QUALITY, 2, 2))

```

```{r}
hist(df$relationship_val)
summary(df$relationship_val)
```

## Difference in Education 

```{r}
hist(df$diff_in_education)
summary(df$diff_in_education)
```

## Difference in Education - Linear Regression


```{r}
summary(lm(relationship_val ~ diff_in_education, data=df))
```

## Absolute Value of Difference in Education


```{r}
df$abs_diff_in_education = abs(df$diff_in_education)
hist(df$abs_diff_in_education)
summary(df$abs_diff_in_education)
```


## Absolute Value of Difference in Education - Linear Regression


```{r}
summary(lm(relationship_val ~ abs_diff_in_education, data=df))
```


## Anova Model on categorical Respondent Years of Education and categorical Partner Years of Education

```{r}
summary(aov(relationship_val ~ factor(RESPONDENT_YRSED) + factor(PARTNER_YRSED), data=df))
summary(aov(relationship_val ~ factor(RESPONDENT_YRSED)*factor(PARTNER_YRSED), data=df))
```

## Anova Model on "Large" Differences in Education vs "Non Large" Differences in Education

```{r}
df$large_diff_in_education <- ifelse(df$abs_diff_in_education >= 4.0, "large", "not large")

summary(aov(relationship_val ~ factor(large_diff_in_education), data=df))
```


## Closer Looker at Couples who have different education levels

```{r}
df_different_education <- df[df$abs_diff_in_education!=0.0,]
hist(df_different_education$abs_diff_in_education)
summary(df_different_education$abs_diff_in_education)
```

```{r}
summary(lm(relationship_val ~ abs_diff_in_education, data=df_different_education))
```






